Reagents: expressing and composing fine-grained concurrency

### 2011-10-17 paper/paper.tex
paper work

Ahmdahl's law tells us that sequential bottlenecks fundamentally limit our
profit from parallelism~\cite{?}.  In practice, the effect is amplified by
another factor: interprocessor communication, often in the form of cache
coherence.  When one thread waits on another, the program pays the cost of
lost parallelism \emph{and} an extra cache miss.  The extra misses can easily
yield parallel \emph{slowdown}, more than negating the benefits of the
remaining parallelism.

@@

@@

In the last section, we saw a range of strategies for communication
and synchronizion at a fine grain:
\begin{itemize}
\item using CAS to make direct, atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents unify these disparate techniques into a single framework for
writing concurrent algorithms.  Some of the techniques are built into
the framework, while others are expressible as reusable abstractions.
% extensibility and composability

@@

@@

\begin{lstlisting}
class MSQueue[A] {
  private final case class Node(
    data: A, next: Ref[Node] = Ref(null)
  )
  private val head = Ref(Node(null))
  private val tail = Ref(head.read!())

  val enq: Reagent[A, Unit] = computed { (x:A) =>
    @tailrec def search: Reagent[Unit,Unit] = 
      tail.read ! () match {
        case    Node(_, r@Ref(null)) => 
          r.cas(null, Node(x))
        case ov@Node(_, Ref(nv)) => 
          tail.cas(ov,nv) !? (); search
      }
    search
  }
  val tryDeq: Reagent[Unit, Option[A]] = head.upd {
    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
    case emp => (emp, None)
  }
}
\end{lstlisting}

@@

@@

>>>>

Ahmdahl's law tells us that sequential bottlenecks fundamentally limit our
profit from parallelism~\cite{?}.  In practice, the effect is amplified by
another factor: interprocessor communication, often in the form of cache
coherence.  When one thread waits on another, the program pays the cost of
lost parallelism \emph{and} an extra cache miss.  The extra misses can easily
accumulate to yield parallel \emph{slowdown}, more than negating the benefits
of the remaining parallelism.

@@

implemented a Scala library

@@

In the last section, we saw a range of strategies for communication
and synchronizion at a fine grain:
\begin{itemize}
\item using CAS to make direct, atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents incorporate these disparate techniques into a uniform framework for
writing concurrent algorithms.  Some of the techniques are built into the
framework, while others are expressible as reusable abstractions.

more easily and concisely
% extensibility and composability

@@

Reagents are a new instance of an old idea: computations can be usefully
represented as data.

% no reason not to present this as an ADT

@@

\begin{lstlisting}
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = Ref(Node(null))
  private val tail = Ref(head.read!())

  @tailrec private def search(x:A): Reagent[Unit,Unit] = 
    tail.read ! () match {
      case Node(_, r@Ref(null)) => 
        r.cas(null, Node(x, new Ref(null)))
      case ov@Node(_, Ref(nv)) => 
        tail.cas(ov,nv) !? (); search(x)
    }
  val enq: Reagent[A, Unit] = computed search
  val tryDeq: Reagent[Unit, Option[A]] = head.upd {
    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
    case emp => (emp, None)
  }
}
\end{lstlisting}

@@

(fractional?) permissions model

@@

We believe it is crucial to have XX, YY together

STM, Communicating transactions and company, elastic transactions and company

CML, Transactional events, join calculus

Paper on composing lock-free data structure operations
